### 2021-02-20 paper/linkagecluster_bk_paper.Rmd
Likelihood and full conditional of linkage cluster model, first attempt

However, if we want to ensure the theoretic purity of each MCMC sample of the linkage structure, we can "recouple" $\mathbf{Z}$ by simply resolving these illegal matchings. Explictly, define clusters $C(i) := \{j | Z_j = i\}$, and define all clusters such that $|C(i)| > 1$ as *conflict clusters*.  Within each cluster, resample the candidate matches according to their Fellegi-Sunter weights. We show this step to be computationally inexpensive and to be a reasonable replication of more organic one-to-one matching. This procedure would be most useful in cases where there are many "near matches" across files, but in most cases, we find this this step to be unnecessary.

@@

The primary contribution of this paper is the establishment of linkage clusters, which allow the reliability parameters $\mathbf{m}^f$, the discernment parameters $\mathbf{u}^f$, and also the linkage probability parameter $\lambda$ to vary according to certain attributes in the record pairs. We shall see that this set up upweights linkage within clusters and downweights linkage across clusters, thus using attribute information in way not yet explored in the literture.

Assign each record in both files to a linkage cluster $S_k \in \{S_1, \ldots, S_K\}$ based on some attribute in the data. We see that the comparison vector

<!-- The classic Fellegi Sunter model operates through weights $w_{ij} = \frac{P(\boldsymbol{\gamma}_{ij}| \text{Records match})}{P(\boldsymbol{\gamma}_{ij}| \text{Records do not match})}$. In our notation, these weights are given as follows: -->
<!-- $$w_{ij} = \frac{P(\boldsymbol{\gamma}_{ij}| Z_j = i)}{P(\boldsymbol{\gamma}_{ij}| Z_j \neq i)} = \frac{\prod_{f=1}^F \prod_{l=1}^{L_f} m_{fl}^{\mathbf{1}_{\gamma_{ij}^f = l}}}{\prod_{f=1}^F \prod_{l=1}^{L_f} u_{fl}^{\mathbf{1}_{\gamma_{ij}^f = l}}}$$ -->

<!-- $$p(\Gamma_{.j}|Z_{j}) = \frac{\mathbf{1}_{z_j < n_A + 1}\lambda\prod_{i=1}^{n_A}\prod_{f=1}^{F}\prod_{l=1}^{L_f} \left(m_{fl}^{\mathbf{1}_{z_j = i}}u_{fl}^{\mathbf{1}_{z_j \neq i}}\right)^{\mathbf{1}_{\gamma_{ij}^f = l}} + \mathbf{1}_{z_j = n_A + 1}(1-\lambda)\prod_{i=1}^{n_A}\prod_{f=1}^{F}\prod_{l=1}^{L_f} u_{fl}^{\mathbf{1}_{\gamma_{ij}^f = l}}}{\sum_{z_j = i}^{n_A}\mathbf{1}_{z_j < n_A + 1}\lambda\prod_{i=1}^{n_A}\prod_{f=1}^{F}\prod_{l=1}^{L_f} \left(m_{fl}^{\mathbf{1}_{z_j = i}}u_{fl}^{\mathbf{1}_{z_j \neq i}}\right)^{\mathbf{1}_{\gamma_{ij}^f = l}} + \mathbf{1}_{z_j = n_A +1}(1-\lambda)\prod_{i=1}^{n_A}\prod_{f=1}^{F}\prod_{l=1}^{L_f} u_{fl}^{\mathbf{1}_{\gamma_{ij}^f = l}}}$$ -->
<!-- We can divide the numerator and denominator by $\prod_{i=1}^{n_A}\prod_{f=1}^{F}\prod_{l=1}^{L_f} u_{fl}^{\mathbf{1}_{\gamma_{ij}^f = l}}$ to obtain -->

<!-- $$p(\Gamma_{.j}|Z_{j}) =\frac{\mathbf{1}_{z_j < n_A + 1}\lambda\prod_{i=1}^{n_A}\prod_{f=1}^{F}\prod_{l=1}^{L_f} \left(\frac{m_{fl}}{u_{fl}}\right)^{\mathbf{1}_{z_j = i} \times \mathbf{1}_{\gamma_{ij}^f = l}} + \mathbf{1}_{z_j = n_A + 1}(1-\lambda)}{\sum_{z_j = 1}^{n_A}\mathbf{1}_{z_j < n_A + 1}\lambda\prod_{i=1}^{n_A}\prod_{f=1}^{F}\prod_{l=1}^{L_f} \left(\frac{m_{fl}}{u_{fl}}\right)^{\mathbf{1}_{z_j = i} \times \mathbf{1}_{\gamma_{ij}^f = l}} + \mathbf{1}_{z_j = n_A + 1}(1-\lambda)}$$ -->

<!-- Our model decouples the draws of the elements of the linkage structure, sampling each $Z_j$ independently from all other $Z_{j'}$. This decoupling allows the incorporation of additional information on each record $j$, thus enabling the linkage cluster formulation proposed in Section XXXX. We clearly admit that this does allow the possibility that $Z_j = i = Z_{j'}$ for some $i\in \mathbf{X}_A$, breaking one-to-one matching requirements in individual Gibbs iterations. However, if our model assumptions are correct and there are no duplicates within files, these sorts of illegal matches are rare and evenly distrubted across records, and thus do not seriously influence our eventual Bayes estimate of the linkage structure. Additionally, these illegal matches can easily be resolved in a post-hoc conflict resolution step, or "recoupling", as described in Section XXXX. -->

>>>>

However, if we want to ensure the theoretic purity of each MCMC sample of the linkage structure, we can "recouple" $\mathbf{Z}$ by simply resolving these illegal matchings. Explictly, define clusters $R(i) := \{j | Z_j = i\}$, and define all clusters such that $|R(i)| > 1$ as *resolution clusters*.  Within each cluster, resample the candidate matches according to their Fellegi-Sunter weights. We show this step to be computationally inexpensive and to be a reasonable replication of more organic one-to-one matching. This procedure would be most useful in cases where there are many "near matches" across files, but in most cases, we find this this step to be unnecessary.

@@

The primary contribution of this paper is the establishment of linkage clusters, which allow the reliability parameters $\mathbf{m}^f$, the discernment parameters $\mathbf{u}^f$, and also the linkage probability parameter $\lambda$ to vary according to certain attributes in the record pairs. We shall see that this set up upweights linkage within clusters and downweights linkage across clusters, thus using attribute information in a way not yet explored in the literture.

Linkage cluster assignment occurs at both at the level of the records and of the record pairs. Assign each record in each file an appropriate linkage cluster $S_k \in \{S_1, \ldots, S_k\}$, denoting the membership record $j$ in the $k^{th}$ cluster by $j \in S_k$. We assign the record pair comparison vector $\gamma_{ij}$ to a cluster only when both records belong to that cluster; otherwise we assign it to $S_0$. For example
$$i \in S_1, j \in S_1 \implies \gamma_{ij} \in S_1$$
$$i \in S_1, j \in S_2 \implies \gamma_{ij} \in S_0$$
Cluster assignments at the record pair level are necessary for the updating the $\mathbf{m}$ and $\mathbf{u}$ parameters. We denote the reliability and discernment parameters assoicated with each field for each cluster as $\mathbf{m}^{fk}$ and $\mathbf{u}^{fk}$ respectively. To improve readability, we modify the definition of $\mathbf{m}_{ij}$ to incorporate the linkage cluster assignment. Explicitly, 
$$\mathbf{m}_{ij} = \prod_{f=1}^{F}\prod_{l=1}^{L_f}\left[ \left(m^{flk}\right)^{\mathbf{1}_{\gamma_{ij} \in S_k}}\left(m^{fl0}\right)^{\mathbf{1}_{\gamma_{ij} \in S_0}}\right]^{\mathbf{1}_{\gamma_{ij}^f = l}}$$
with an analogous definition for $\mathbf{u}_{ij}$.

Additionally, define the vector of linkage probabilities $\Lambda_k = (\lambda_{k1}, \lambda_{k2}, \lambda_{k3})$, where $\lambda_{k1} = P(z_{j\in S_k} \in S_k)$ is the probability that a record in $\mathbf{X}_{B_k}$ has a match within its cluster in $\mathbf{X}_{A_k}$, $\lambda_{k2} = P(z_{j \in S_k} \in S_{k'})$ is the probability that such a record $\mathbf{X}_{B_k}$ has a match outside of its cluster $\mathbf{X}_{A_{k'}}$ for some $k' \neq k$, and $\lambda_{k3} = P(z_{j\in S_k} = n_A + 1)$ is the probability of a non-match. This is a multinomial distribution where we necessarily have $\lambda_{k3} = 1 - \lambda_{k1} - \lambda_{k2}$. These parameters use linkage cluster assignment at the record level.

Our new model specification becomes

$$P(\Gamma|\mathbf{Z}, \mathbf{m}, \mathbf{u}, \boldsymbol{\theta}) =\prod_{k=1}^K \prod_{j \in \mathbf{X}_B^k}  \left[\lambda_{k1}^{\mathbf{1}_{z_j \in S_k}}\lambda_{k2}^{\mathbf{1}_{z_j \in S_{k'}}} \lambda_{k3}^{\mathbf{1}_{z_j = n_A + 1}}\prod_{i=1}^{n_A}\mathbf{m}_{ij}^{\mathbf{1}_{z_j = i}}\mathbf{u}_{ij}^{\mathbf{1}_{z_j \neq i}}\right]$$

$$\mathbf{m^{fk}} \sim \text{Dirichlet}(\alpha^{f1k}, \ldots, \alpha^{fL_fk})$$
$$\mathbf{u^{fk}} \sim \text{Dirichlet}(\beta^{f1k}, \ldots, \beta^{fL_fk})$$
$$Z_{j \in S_k} | \Lambda_k =
\begin{cases} 
    \frac{1}{n_{A_k}}\lambda_{k1}  & z_j \in S_k; \\
    \frac{1}{n_A - n_{A_k}}\lambda_{k2}  & z_j \in S_{k'}; \\
     \lambda_{k3} & z_j = n_A + 1 \\
\end{cases}$$

$$\Lambda \sim \text{Dirichlet}(\Lambda_0)$$
where $\Lambda_0 = (\Lambda_{k1}, \Lambda_{k2}, \Lambda_{k3}).$

**THIS WOULD GO IN APPENDIX**: Through similar derivations, this yields full conditionals:
The $\mathbf{m}$ and $\mathbf{u}$ parameters are updated through standard multinomial-dirichlet mechanics. Thus we have 
$$\mathbf{m^{fk}}|\mathbf{Z}, \Gamma \sim \text{Dirichlet}(\alpha^{f1k}(\mathbf{Z}), \ldots, \alpha^{fL_fk}(\mathbf{Z}))$$

$$\mathbf{u^{fk}}|\mathbf{Z}, \Gamma \sim \text{Dirichlet}(\beta^{f1k}(\mathbf{Z}), \ldots, \beta^{fL_fk}(\mathbf{Z}))$$
where $$\alpha_{flk}(\mathbf{Z})= \sum_{i,j, k} \mathbf{1}_{obs(\gamma_{ij}^f)}\mathbf{1}_{\gamma_{ij}^f = l} \mathbf{1}_{z_j = i}\mathbf{1}_{\gamma_{ij} \in S_k}$$

$$\text{and } \beta_{flk}(\mathbf{Z})= \sum_{i,j, k} \mathbf{1}_{obs(\gamma_{ij}^f)}\mathbf{1}_{\gamma_{ij}^f = l} \mathbf{1}_{z_j \neq i}\mathbf{1}_{\gamma_{ij}^f = l}$$

We also have straightforward posterior distritutions for the linkage probability vector $\Lambda_k$:

$$\Lambda_k | \mathbf{Z} \sim \text{Dirichlet}(\mathbf{D_k} + \Lambda_0)$$

where $\mathbf{D}_k = (D_{k1}, D_{k2}, D_{k3})$ with 
$$D_{k1} = \sum_{j \in \mathbf{X}_{B_k}} \mathbf{1}_{z_j \in S_k}$$ $$D_{k2} = \sum_{j \in \mathbf{X}_{B_k}} \mathbf{1}_{z_j \in S_{k'}}$$ $$\text{and } D_{k3} = \sum_{j \in \mathbf{X}_{B_k}} \mathbf{1}_{z_j = n_A + 1}$$

Our full conditional for $Z_j$ follows a familiar form:

$$Z_{j \in S_k}^{(s+1)} | \mathbf{m}^k, \mathbf{u}^k, \Gamma, \Lambda_k =
\begin{cases} 
    \frac{\lambda_{k1}}{n_{A_k}} w_{ij} & z_j \in S_k; \\
    \frac{\lambda_{k2}}{n_A - n_{A_k}} w_{ij}  & z_j \in S_{k'}; \\
     \lambda_{k3} & z_j = n_A + 1 \\
\end{cases}$$

And after integrating over $\Lambda_k$ and rearraning terms,

$$Z_{j \in S_k}^{(s+1)} | \mathbf{m}^k, \mathbf{u}^k, \Gamma,  =
\begin{cases} 
    \frac{D_{k1} + \Lambda_{01}}{n_{A_k}(n_B - D_{k1} - D_{k2} + \Lambda_{03})} w_{ij} & z_j \in S_k; \\
    \frac{D_{k2} + \Lambda_{02}}{(n_A - n_{A_k})(n_B - D_{k1} - D_{k2} + \Lambda_{03})} w_{ij}  & z_j \in S_{k'}; \\
     1 & z_j = n_A + 1 \\
\end{cases}$$

<!-- While the base model tracked matches and non-matches, our linkage cluster requires us additionally to be able to track matches within and across linkage clusters. Define the match parameter vector $\boldsymbol{\lambda}^k = (\lambda^{k1}, \lambda^{k0}, 1 - \lambda^{k1} - \lambda^{k0})$ where $\lambda^{k1}$ is the probability of matching within cluster $k$ and $\lambda^{k0}$ is probability of matching across cluster. (Man this is hard to write explicitly! Will need to revise). -->

<!-- Associate each record in both files with linkage cluster $k \in \{1, \ldots, K\}$, and represent the association of record $j$ in with cluster $k$ by saying $C(j) = k$. Denote the records in $\mathbf{X}_A$ and $\mathbf{X}_B$ associated with cluster $k$ by $\mathbf{X}_A^k$ and $\mathbf{X}_B^k$ respectively. We assign record pairs to cluster $k$ only if both records are assigned to that cluster, ie when $C(i) = k = C(j)$. Otherwise, the pair is assigned to cluster $0$. -->

### 2021-02-20 paper/linkagecluster_bk_paper.tex
Likelihood and full conditional of linkage cluster model, first attempt

However, if we want to ensure the theoretic purity of each MCMC sample
of the linkage structure, we can ``recouple'' \(\mathbf{Z}\) by simply
resolving these illegal matchings. Explictly, define clusters
\(C(i) := \{j | Z_j = i\}\), and define all clusters such that
\(|C(i)| > 1\) as \emph{conflict clusters}. Within each cluster,
resample the candidate matches according to their Fellegi-Sunter
weights. We show this step to be computationally inexpensive and to be a
reasonable replication of more organic one-to-one matching. This
procedure would be most useful in cases where there are many ``near
matches'' across files, but in most cases, we find this this step to be
unnecessary.

@@

The primary contribution of this paper is the establishment of linkage
clusters, which allow the reliability parameters \(\mathbf{m}^f\), the
discernment parameters \(\mathbf{u}^f\), and also the linkage
probability parameter \(\lambda\) to vary according to certain
attributes in the record pairs. We shall see that this set up upweights
linkage within clusters and downweights linkage across clusters, thus
using attribute information in way not yet explored in the literture.

Assign each record in both files to a linkage cluster
\(S_k \in \{S_1, \ldots, S_K\}\) based on some attribute in the data. We
see that the comparison vector

>>>>

However, if we want to ensure the theoretic purity of each MCMC sample
of the linkage structure, we can ``recouple'' \(\mathbf{Z}\) by simply
resolving these illegal matchings. Explictly, define clusters
\(R(i) := \{j | Z_j = i\}\), and define all clusters such that
\(|R(i)| > 1\) as \emph{resolution clusters}. Within each cluster,
resample the candidate matches according to their Fellegi-Sunter
weights. We show this step to be computationally inexpensive and to be a
reasonable replication of more organic one-to-one matching. This
procedure would be most useful in cases where there are many ``near
matches'' across files, but in most cases, we find this this step to be
unnecessary.

@@

The primary contribution of this paper is the establishment of linkage
clusters, which allow the reliability parameters \(\mathbf{m}^f\), the
discernment parameters \(\mathbf{u}^f\), and also the linkage
probability parameter \(\lambda\) to vary according to certain
attributes in the record pairs. We shall see that this set up upweights
linkage within clusters and downweights linkage across clusters, thus
using attribute information in a way not yet explored in the literture.

Linkage cluster assignment occurs at both at the level of the records
and of the record pairs. Assign each record in each file an appropriate
linkage cluster \(S_k \in \{S_1, \ldots, S_k\}\), denoting the
membership record \(j\) in the \(k^{th}\) cluster by \(j \in S_k\). We
assign the record pair comparison vector \(\gamma_{ij}\) to a cluster
only when both records belong to that cluster; otherwise we assign it to
\(S_0\). For example
\[i \in S_1, j \in S_1 \implies \gamma_{ij} \in S_1\]
\[i \in S_1, j \in S_2 \implies \gamma_{ij} \in S_0\] Cluster
assignments at the record pair level are necessary for the updating the
\(\mathbf{m}\) and \(\mathbf{u}\) parameters. We denote the reliability
and discernment parameters assoicated with each field for each cluster
as \(\mathbf{m}^{fk}\) and \(\mathbf{u}^{fk}\) respectively. To improve
readability, we modify the definition of \(\mathbf{m}_{ij}\) to
incorporate the linkage cluster assignment. Explicitly,
\[\mathbf{m}_{ij} = \prod_{f=1}^{F}\prod_{l=1}^{L_f}\left[ \left(m^{flk}\right)^{\mathbf{1}_{\gamma_{ij} \in S_k}}\left(m^{fl0}\right)^{\mathbf{1}_{\gamma_{ij} \in S_0}}\right]^{\mathbf{1}_{\gamma_{ij}^f = l}}\]
with an analogous definition for \(\mathbf{u}_{ij}\).

Additionally, define the vector of linkage probabilities
\(\Lambda_k = (\lambda_{k1}, \lambda_{k2}, \lambda_{k3})\), where
\(\lambda_{k1} = P(z_{j\in S_k} \in S_k)\) is the probability that a
record in \(\mathbf{X}_{B_k}\) has a match within its cluster in
\(\mathbf{X}_{A_k}\), \(\lambda_{k2} = P(z_{j \in S_k} \in S_{k'})\) is
the probability that such a record \(\mathbf{X}_{B_k}\) has a match
outside of its cluster \(\mathbf{X}_{A_{k'}}\) for some \(k' \neq k\),
and \(\lambda_{k3} = P(z_{j\in S_k} = n_A + 1)\) is the probability of a
non-match. This is a multinomial distribution where we necessarily have
\(\lambda_{k3} = 1 - \lambda_{k1} - \lambda_{k2}\). These parameters use
linkage cluster assignment at the record level.

(THIS WOULD GO IN APPENDIX) Our new model specification becomes

\[P(\Gamma|\mathbf{Z}, \mathbf{m}, \mathbf{u}, \boldsymbol{\theta}) =\prod_{k=1}^K \prod_{j \in \mathbf{X}_B^k}  \left[\lambda_{k1}^{\mathbf{1}_{z_j \in S_k}}\lambda_{k2}^{\mathbf{1}_{z_j \in S_{k'}}} \lambda_{k3}^{\mathbf{1}_{z_j = n_A + 1}}\prod_{i=1}^{n_A}\mathbf{m}_{ij}^{\mathbf{1}_{z_j = i}}\mathbf{u}_{ij}^{\mathbf{1}_{z_j \neq i}}\right]\]

\[\mathbf{m^{fk}} \sim \text{Dirichlet}(\alpha^{f1k}, \ldots, \alpha^{fL_fk})\]
\[\mathbf{u^{fk}} \sim \text{Dirichlet}(\beta^{f1k}, \ldots, \beta^{fL_fk})\]
\[Z_{j \in S_k} | \Lambda_k =
\begin{cases} 
    \frac{1}{n_{A_k}}\lambda_{k1}  & z_j \in S_k; \\
    \frac{1}{n_A - n_{A_k}}\lambda_{k2}  & z_j \in S_{k'}; \\
     \lambda_{k3} & z_j = n_A + 1 \\
\end{cases}\]

\[\Lambda \sim \text{Dirichlet}(\Lambda_0)\] where
\(\Lambda_0 = (\Lambda_{k1}, \Lambda_{k2}, \Lambda_{k3}).\)

\textbf{THIS WOULD GO IN APPENDIX}: Through similar derivations, this
yields full conditionals: The \(\mathbf{m}\) and \(\mathbf{u}\)
parameters are updated through standard multinomial-dirichlet mechanics.
Thus we have
\[\mathbf{m^{fk}}|\mathbf{Z}, \Gamma \sim \text{Dirichlet}(\alpha^{f1k}(\mathbf{Z}), \ldots, \alpha^{fL_fk}(\mathbf{Z}))\]

\[\mathbf{u^{fk}}|\mathbf{Z}, \Gamma \sim \text{Dirichlet}(\beta^{f1k}(\mathbf{Z}), \ldots, \beta^{fL_fk}(\mathbf{Z}))\]
where
\[\alpha_{flk}(\mathbf{Z})= \sum_{i,j, k} \mathbf{1}_{obs(\gamma_{ij}^f)}\mathbf{1}_{\gamma_{ij}^f = l} \mathbf{1}_{z_j = i}\mathbf{1}_{\gamma_{ij} \in S_k}\]

\[\text{and } \beta_{flk}(\mathbf{Z})= \sum_{i,j, k} \mathbf{1}_{obs(\gamma_{ij}^f)}\mathbf{1}_{\gamma_{ij}^f = l} \mathbf{1}_{z_j \neq i}\mathbf{1}_{\gamma_{ij}^f = l}\]

We also have straightforward posterior distritutions for the linkage
probability vector \(\Lambda_k\):

\[\Lambda_k | \mathbf{Z} \sim \text{Dirichlet}(\mathbf{D_k} + \Lambda_0)\]

where \(\mathbf{D}_k = (D_{k1}, D_{k2}, D_{k3})\) with
\[D_{k1} = \sum_{j \in \mathbf{X}_{B_k}} \mathbf{1}_{z_j \in S_k}\]
\[D_{k2} = \sum_{j \in \mathbf{X}_{B_k}} \mathbf{1}_{z_j \in S_{k'}}\]
\[\text{and } D_{k3} = \sum_{j \in \mathbf{X}_{B_k}} \mathbf{1}_{z_j = n_A + 1}\]

Our full conditional for \(Z_j\) follows a familiar form:

\[Z_{j \in S_k}^{(s+1)} | \mathbf{m}^k, \mathbf{u}^k, \Gamma, \Lambda_k =
\begin{cases} 
    \frac{\lambda_{k1}}{n_{A_k}} w_{ij} & z_j \in S_k; \\
    \frac{\lambda_{k2}}{n_A - n_{A_k}} w_{ij}  & z_j \in S_{k'}; \\
     \lambda_{k3} & z_j = n_A + 1 \\
\end{cases}\]

And after integrating over \(\Lambda_k\) and rearraning terms,

\[Z_{j \in S_k}^{(s+1)} | \mathbf{m}^k, \mathbf{u}^k, \Gamma,  =
\begin{cases} 
    \frac{D_{k1} + \Lambda_{01}}{n_{A_k}(n_B - D_{k1} - D_{k2} + \Lambda_{03})} w_{ij} & z_j \in S_k; \\
    \frac{D_{k2} + \Lambda_{02}}{(n_A - n_{A_k})(n_B - D_{k1} - D_{k2} + \Lambda_{03})} w_{ij}  & z_j \in S_{k'}; \\
     1 & z_j = n_A + 1 \\
\end{cases}\]
